Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: A Statistical Search Through Palomar Transient Factory Data

### 2013-06-20 text/paper/ms.tex
paper update

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{natbib}
\usepackage{longtable, hyperref, graphicx, subfigure, epstopdf, multirow, setspace, verbatim, lmodern} 
\usepackage{aas_macros}

@@

\title{Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case
 of the Palomar Transient Factory}
\author{Adrian~M.~Price-Whelan\altaffilmark{1,10} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda Fournier\altaffilmark{2}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{8}, Jason Surace\altaffilmark{9}}
%, Joshua S.\ Bloom\altaffilmark{\ref{cal}}, S.\ Bradley Cenko\altaffilmark{\ref{cal}}, Mansi M.\ Kasliwal\altaffilmark{\ref{calt}}, Shrinivas R.\ Kulkarni\altaffilmark{\ref{calt}}, Nicholas~M.~Law\altaffilmark{\ref{to},\ref{dun}}, Peter Nugent\altaffilmark{\ref{lbnl}}, Dovi Poznanski\altaffilmark{\ref{cal},\ref{calt},\ref{eins}}, Robert M.\ Quimby\altaffilmark{\ref{calt}}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Cahill Center for Astrophysics, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{6}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}
\altaffiltext{8}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{9}{Infrared Processing and Analysis Center, California Institute of Technology, Mail Stop 100-22, Pasadena, CA 91125, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{cal} Department of Astronomy, University of California, Berkeley, CA  94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{to} Dunlap Institute for Astronomy and Astrophysics, University of Toronto, 50 St.\ George Street, Toronto M5S 3H4, Ontario, Canada}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{lbnl} Computational Cosmology Center, Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{dun} Dunlap Fellow}
\altaffiltext{10}{NSF Graduate Fellow}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations, a total of $1.1\times10^9$ light curves, to uncover three candidate sparse-field microlensing events. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large datasets, both of which will be useful to future wide-field, time-domain surveys such as that planned with the Large Synoptic Survey Telescope.

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and, most recently, to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects such as low-mass stars, sub-stellar objects, and isolated neutron stars.

%The microlensing event rate for a given region of the sky depends on the space and velocity distributions of the sources and on the space, velocity, and mass distribution of the lenses.
To maximize the event rate, microlensing surveys typically focus on high-density stellar regions such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.][]{original_ogle, original_macho, eros_original, crotts1996}. Of course, microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the event rate in low-density stellar regions will certainly be lower than that near e.g., the Bulge \citep[][]{ogle_optical_depth, macho_optical_depth, wood_optical_depth, eros_optical_depth}, these events are expected to involve closer sources and lenses, and less crowded backgrounds, compared to those found by modern microlensing surveys \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and therefore mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii therefore tend to be larger (a few milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would therefore be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. We examine the selection methods used by microlensing surveys to identify event signatures in light curves and an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search} we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}.

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF is an automated transient detection system \citep{nick2009,rau2009}. Data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48 inch Oschin Schmidt telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2010}. The data are processed in real time by a pipeline that identifies transients of interest and passes these to a dedicated photometric follow-up telescope; this has proved an excellent tool for e.g., early detection of supernovae \citep{nugent2011}.

Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al., in prep). Catalogs are then generated using SExtractor for source identification \citep{bertin96}. The absolute photometric calibration of the images is good to a systematic limit of $\sim2\%$ for photometric nights, as described in \cite{ofek2012}.  The final light curves are produced using relative photometric calibration, which refines the calibration to $<$1\% for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{levitan2011, ofek2012}. The resulting archive is an excellent resource for the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}.

@@

As of 2012 Dec, the PTF footprint includes $\sim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\sim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint})\footnote{For an interactive visualization of the PTF survey footprint, see \url{http://bit.ly/PTFSky}.}. The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, what time of the year it is visible, and how high a priority it is given by the scheduler. The imaging cadences range from 1-5 days, but there is a higher-cadence campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period.

@@

What would a microlensing event look like in a typical PTF light curve? If the source is an unblended point source and the lens is a foreground, dim object, a microlensing event is fully described by three parameters: the angular impact parameter $u_0$, the peak time of the event $t_0$, and the timescale of the event (Einstein crossing time) $t_E$. The amplification factor $A$ and flux $F$ can then be defined as a function of time: 
\begin{align}
	u(t) &= \sqrt{u_0^2 + 2\Big(\frac{t-t_0}{t_E}\Big)},\\
	A(t) &= \frac{u^2 + 2}{u\sqrt{u^2 + 4}},\\
	F(t) &= A(t)\times F_{source}.
\end{align}
The microlensing perturbation can also be expressed as:
\begin{align}\label{eq:ml_model}
	m(t) &= m_0(t) - 2.5\log A(t),
\end{align}
where $m_0(t)$ is the unperturbed, but possibly time-variable, magnitude of the source.

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$, but different $u_0$, on a random well-sampled PTF light curve.

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. Light curves of any transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive datapoints brighter than some threshold,
	\item compared to flat or linear light-curve models, a microlensing model best describe the data, as determined by using a $\Delta\chi^2$ test, and
	\item the microlensing model event parameters have physically reasonable values.
\end{enumerate}
%Difference imaging is ideal for studying crowded fields where photometric measurements are difficult. It also limits the number of light curves that must be fit with a model, thus saving on computation time and allowing for real-time detection of events, so that high-cadence follow-up observations of interesting sources can be triggered.

@@

\subsection{The Variability Indices} %\label{sec:event_selection}
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$\footnote{We do not to implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.}. $\sigma/\mu$ is the ratio of the root-variance to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 8 photometry \citep{DR8paper} to identify candidate quasars \citep[using the cuts described in][]{richards02} and examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by e.g., supernovae.

We visually inspect each of the remaining $\sim$2000 light curves. We identify $\sim$1100 quasars that did not meet the \cite{richards02} color cut, or objects with unknown long-term variability (e.g., Mira-type variables). An additional $\sim$600 light curves have bad or poorly calibrated data that mimics a transient increase in flux: diffraction spikes, ghosts, and scattered light can cause non-Gaussian, seeing-dependent variability. Where appropriate, as an additional test, we examine the PTF images to verify the data quality. A small fraction of the selected light curves are variable stars that made it past our periodic variability cut.

We are left with $\sim$300 unclassified transients. Through visual inspection, we classify the bulk of these as novae, supernovae, flares, or outbursting systems, and identify 11 microlensing event candidates. We then search the literature for additional photometric data for these 11 objects. Most have a counterpart in WISE \citep{WISE}, while a few have a counterpart in 2MASS \citep{cutri03}; see Table~\ref{table:photometry} for a summary. We also obtained spectroscopy for several of these candidates. Below we detail what can be learned from these data. Generally, additional deeper optical/infrared imaging and/or optical spectroscopy is required to draw firm conclusions about the nature of the object.

To summarize: we rule out five of the 11 candidates as being likely extragalactic transients based on photometry from WISE and/or 2MASS. Three of the remaining objects are ambiguous because they are not detected in WISE and/or 2MASS; follow-up imaging or spectroscopy is needed to draw any conclusions about these candidates. The remaining three objects (PTF1J061800.25$+$203142.5, 
J161502.39$+$540053.8, and J172826.08$+$692501.1) are likely stars and the PTF-detected variability plausible microlensing events. We emphasize, however, that lacking simultaneous, multi-color photometry during the events, we cannot claim these as microlensing events with any confidence. %The reader may wish to focus on the description of the available data for these three candidates before moving on to the next section.

@@

We observed PTF1J0618 with the Double Spectrograph \citep[DBSP;][]{oke_gunn} on the Hale 5-m telescope at Palomar Observatory, CA, on 2013 Feb 19; see Figure~\ref{fig:1206i_spectrum}. The spectrum was obtained with the D55 dichroic; from the atmospheric cutoff to 5500 \AA, the grating had 600 line~mm$^{-1}$ and was blazed at 4000 \AA, giving a resolution of 1.1 \AA. From 6300-8800 \AA, the grating was 158 line~mm$^{-1}$, was blazed at 7500\ \AA, and gave a resolution was 2.5 \AA. The data reduction was performed by S.~Tang using standard IRAF routines\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.}.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=.48\textwidth, trim=17 15 15 40, clip] {figures/PTFS1206i_spectrum515.pdf}
\caption{\label{fig:1206i_spectrum} DBSP spectrum of PTF1J0618 smoothed with a Gaussian filter ($\sigma = 1$). This object is most likely a M2 dwarf star. The timescale of the event detected by PTF is inconsistent with those generally observed in flares, which are the mostly likely contaminants in identifying microlensing events involving M stars.}
\end{figure}

>>>>

\usepackage{amssymb,amsmath}
\usepackage{amsfonts}
\usepackage{natbib}
\usepackage{longtable, graphicx, subfigure, epstopdf, multirow, setspace, verbatim, lmodern} 
\usepackage{aas_macros}
\newcounter{address}
\setcounter{address}{1}

@@

\title{Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: A Statistical Search Through Palomar Transient Factory Data}
\author{Adrian~M.~Price-Whelan\altaffilmark{1,8} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda P. Fournier\altaffilmark{2,3}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{6}, Jason Surace\altaffilmark{7}}
%, Joshua S.\ Bloom\altaffilmark{\ref{cal}}, S.\ Bradley Cenko\altaffilmark{\ref{cal}}, Mansi M.\ Kasliwal\altaffilmark{\ref{calt}}, Shrinivas R.\ Kulkarni\altaffilmark{\ref{calt}}, Nicholas~M.~Law\altaffilmark{\ref{to},\ref{dun}}, Peter Nugent\altaffilmark{\ref{lbnl}}, Dovi Poznanski\altaffilmark{\ref{cal},\ref{calt},\ref{eins}}, Robert M.\ Quimby\altaffilmark{\ref{calt}}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{6}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}
%\altaffiltext{9}{Infrared Processing and Analysis Center, California Institute of Technology, Mail Stop 100-22, Pasadena, CA 91125, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{cal} Department of Astronomy, University of California, Berkeley, CA  94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{to} Dunlap Institute for Astronomy and Astrophysics, University of Toronto, 50 St.\ George Street, Toronto M5S 3H4, Ontario, Canada}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{lbnl} Computational Cosmology Center, Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{dun} Dunlap Fellow}
\altaffiltext{8}{NSF Graduate Fellow}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations, a total of $1.1\times10^9$ light curves, to uncover three candidate sparse-field microlensing events. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope.

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and, most recently, to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

%The microlensing event rate for a given region of the sky depends on the space and velocity distributions of the sources and on the space, velocity, and mass distribution of the lenses.
To maximize the event rate, microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.][]{original_ogle, original_macho, eros_original, crotts1996}. Of course, microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the event rate in low-density stellar regions will certainly be lower than that near, e.g., the Bulge \citep[][]{ogle_optical_depth, macho_optical_depth, wood_optical_depth, eros_optical_depth}, these events are expected to involve closer sources and lenses, and less crowded backgrounds, compared to those found by modern microlensing surveys \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and, therefore, mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii, therefore, tend to be larger (a few milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would, therefore, be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, which is described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. A simple extrapolation from previous all-sky rate estimates suggests $\sim$1--10 detectable microlensing events within the PTF footprint over the three year survey \citep{han2008, gaudi2008}. We examine the selection methods used by microlensing surveys to identify event signatures in light curves and an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search}, we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}. A detailed investigation of the all-sky microlensing event rate,  generated from a model that includes realistic mass and velocity distributions, will be presented in Fournier et al. (in prep).

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48-nch Oschin Schmidt Telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (2.2$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2009,rau2009,nick2010}. %The data are processed in real time by a pipeline that i

While a real-time image-differencing pipeline identifies transients of interest and passes these to a dedicated photometric follow-up telescope, a separate pipeline generates PTF light curves. Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al., in prep). Catalogs are produced using SExtractor for source identification \citep{bertin96}. The absolute photometric calibration of the images is good to a systematic limit of $\sim$2\% for photometric nights, as described in \cite{ofek2012}.  The final light curves are produced using relative photometric calibration, which refines the calibration to $<$1\% for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{levitan2011, ofek2012}. The resulting archive is an excellent resource for the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}. %A separate pipeline identifies transients of interest and passes these to a dedicated photometric follow-up telescope; this has proved an excellent tool for e.g., early detection of supernovae \citep{nugent2011}.

@@

As of 2012 Dec, the PTF footprint includes $\sim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\sim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint}). The PTF survey footprint is not uniformly sampled either spatially or temporally.\footnote{For an interactive visualization of the PTF survey footprint, see \url{http://bit.ly/PTFSky}.} Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, what time of the year it is visible, and how high a priority it is given by the scheduler. The imaging cadences range from 1-5 days, but there is a higher-cadence campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period.

@@

What would a microlensing event look like in a typical PTF light curve? If the source is an unblended point source and the lens is a foreground, dim object, a microlensing event is fully described by three parameters: the angular impact parameter $u_0$, the peak time of the event $t_0$, and the timescale of the event (Einstein crossing time) $t_E$. In terms of a dimensionless projected distance between the source and lens, $u$, the amplification factor $A$ and flux $F$ as a function of time can be defined as: 
\begin{align}
	u(t) &= \sqrt{u_0^2 + 2\Big(\frac{t-t_0}{t_E}\Big)},\\
	A(t) &= \frac{u^2 + 2}{u\sqrt{u^2 + 4}},\\
	F(t) &= A(t)\times F_{source}.
\end{align}
\citep{paczynski1986}. The microlensing perturbation can also be expressed as:
\begin{align}\label{eq:ml_model}
	m(t) &= m_0(t) - 2.5\log A(t),
\end{align}
where $m_0(t)$ is the unperturbed, but possibly time-variable, magnitude of the source.

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$, but different angular impact parameter, $u_0$, on a random well-sampled PTF light curve.

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. The light curves of transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from, e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive data points brighter than some threshold,
	\item compared to flat or linear light-curve models, a microlensing model best describe the data, as determined by using a $\Delta\chi^2$ test, and
	\item the microlensing model event parameters have physically reasonable values.
\end{enumerate}
%Difference imaging is ideal for studying crowded fields where photometric measurements are difficult. It also limits the number of light curves that must be fit with a model, thus saving on computation time and allowing for real-time detection of events, so that high-cadence follow-up observations of interesting sources can be triggered.

@@

\subsection{The Variability Indices} %\label{sec:event_selection}
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$\footnote{We do not implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.}. $\sigma/\mu$ is the ratio of the sample standard deviation to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 7 photometry \citep{DR7paper} to identify candidate quasars \citep[using the cuts described in][]{richards02} and examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by e.g., supernovae.

We visually inspect each of the remaining $\sim$2000 light curves. We identify an additional $\sim$1100 objects with unknown long-term variability (e.g, Mira-type variables, quasars not in the SDSS footprint). An additional $\sim$600 light curves have bad or poorly calibrated data that mimics a transient increase in flux: diffraction spikes, ghosts, and scattered light can cause non-Gaussian, seeing-dependent variability. Where appropriate, as an additional test, we examine the PTF images to verify the data quality. A small fraction of the selected light curves are variable stars that made it past our periodic variability cut.

We are left with $\sim$300 unclassified transients. Through visual inspection of these light curves, we classify the bulk of these as novae, supernovae, flares, or outbursting systems, and identify 11 microlensing event candidates. We then search the literature for additional photometric data for these 11 objects. Most have a counterpart in WISE \citep{WISE}, while a few have a counterpart in 2MASS \citep{cutri03}; see Table~\ref{table:photometry} for a summary. We also obtained spectroscopy for several of these candidates. Below we detail what can be learned from these data. Generally, additional deeper optical/infrared imaging and/or optical spectroscopy is required to draw firm conclusions about the nature of the object.

To summarize: we rule out five of the 11 candidates as being likely extragalactic transients based on photometry from WISE and/or 2MASS. Three of the remaining objects are ambiguous because they are not detected in WISE and/or 2MASS; follow-up imaging or spectroscopy is needed to draw any conclusions about these candidates. The remaining three objects (PTF1J061800.25$+$203142.5, 
J161502.39$+$540053.8, and J172826.08$+$692501.1) are likely stars and, given the observed PTF variability, plausible microlensing events. We emphasize, however, that lacking simultaneous, multi-color photometry during the events, we cannot claim these as microlensing events with any confidence. %The reader may wish to focus on the description of the available data for these three candidates before moving on to the next section.

@@

We observed PTF1J0618 with the Double Spectrograph \citep[DBSP;][]{oke_gunn} on the Hale 5-m telescope at Palomar Observatory, CA, on 2013 Feb 19; see Figure~\ref{fig:1206i_spectrum}. The spectrum was obtained with the D55 dichroic; from the atmospheric cutoff to 5500 \AA, the grating had 600 line~mm$^{-1}$ and was blazed at 4000 \AA, giving a resolution of 1.1 \AA. From 6300-8800 \AA, the grating was 158 line~mm$^{-1}$, was blazed at 7500\ \AA, and gave a resolution of 2.5 \AA. The data were reduced using standard IRAF routines\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.}.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=.48\textwidth, trim=17 15 15 40, clip] {figures/PTFS1206i_new.pdf}
\caption{\label{fig:1206i_spectrum} DBSP spectrum of PTF1J0618 smoothed with a Gaussian filter ($\sigma = 1$). This object is most likely a M2 dwarf star. The timescale of the event detected by PTF is inconsistent with those generally observed in flares, which are the mostly likely contaminants in identifying microlensing events involving M stars.}
\end{figure}
